---
title: "Stat_652"
author: "Ria Gupta"
date: "06/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading flight data

The data are on flights from three New York City airports 
in 2013, from the `nycflights13` package. 
Data were combined from four datasets from this package:

```{r}
library(tidyverse)
library(nycflights13)
#help(flights)
#help(weather)
#help(airports)
#help(planes)
fltrain <- read.csv("C:\\Users\\Karan\\Downloads\\stat\\project\\fltrain.csv.gz")
```


```{r}
dim(fltrain)

```

There are 43 variables measured on 200,000 flights. 


##Converting character to factor variables
```{r}
fl <- fltrain
for(i in 1:ncol(fl)) {
  if(typeof(fl[[i]]) == "character") {
    fl[[i]] <- factor(fl[[i]])
  }
}

```
## Missing data


### Calculating percentage of missing values for each column

```{r}
p<-function(x) {sum(is.na(x))/length(x)*100}
apply(fl,2,p)

```
### Action:
From above missing values percentage we observe, that predictors such as wind_gust, pressure, type, manufacturer, model,engines,seats, engine, speed have more than 5% values missing. From thumb of rule, we remove these variables.

```{r}

fl <- fl%>% 
  select(-type,-manufacturer,-model,-engines,-seats, -speed, -engine,-wind_gust,-pressure,-year.y)

```

###to omit the rows with missing data

```{r}

fl <- na.omit(fl)
summary(fl)

```

### Summaries of the response variable `dep_delay`

The departure delays variable is highly right-skewed.

```{r}

range(fl$dep_delay)
fivenum(fl$dep_delay)
quantile(fl$dep_delay,probs = c(0.01,0.05,0.1,0.25,.5,.75,.90,.95,.99))
mean(fl$dep_delay >= 60) # about 15,000 or 8% of flights

```

###Summaries of departure delay by NYC airport:
```{r}

Q3 <- function(x) { quantile(x,probs=.75) }
fl %>% group_by(origin) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 

```

###Summaries of departure delay by airline (carrier).

```{r}

fl %>% group_by(carrier) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 
fl %>% group_by(origin,carrier) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 
fl %>% group_by(dest,carrier) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 

```

###Summaries of departure delay by date:
  
```{r}

fl %>% group_by(month,day) %>% 
  summarize(n=n(),med_d = mean(dep_delay),max_d = max(dep_delay)) %>% 
  arrange(desc(med_d)) %>% head(10) 

```


###Summaries of departure delay by precipitation:
```{r}

fl %>% mutate(haveprecip = factor(precip>0)) %>% group_by(haveprecip) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(med_d)) %>% head(10) 

```

### Ranking and scaling dep_delay

```{r}

den <- nrow(fl)+1
fl <- fl %>% mutate(dep_delay = rank(dep_delay)/den)
ggplot(fl,aes(x=dep_delay)) + geom_histogram(binwidth=.01)

```

###data wrangling

```{r}

library(lubridate)
fl <- fl %>% 
  mutate(dep_date = make_date(year.x,month,day)) %>% 
  select(-year.x,-month,-day,-dep_time,-arr_time,-arr_delay,
         -sched_arr_time,-tailnum,-flight,-name,-air_time,
         -hour,-minute,-time_hour,-tz,-dst,-dest) %>%
  mutate(precip = as.numeric(precip>0))
fl <- mutate(fl,logalt = log(alt)) %>% select(-alt)
fl <- mutate(fl,logdistance = log(distance)) %>% select(-distance)

```


## Associations between `dep_delay` and quantitative predictors


```{r,eval=FALSE}

ggplot(fl,aes(x=dep_date,y=dep_delay)) + geom_point(alpha=.01) + geom_smooth()

ggplot(fl,aes(x=sched_dep_time,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()


ggplot(fl,aes(x=log(distance),y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()


ggplot(fl,aes(x=temp,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()

ggplot(fl,aes(x=dewp,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()

```


## Split training set in two for tuning

```{r}

set.seed(123)
tr_size <- ceiling(2*nrow(fl)/3)
train <- sample(1:nrow(fl),size=tr_size)
fl_tr <- fl[train,]
fl_te <- fl[-train,]

```


## Learning methods


###Random Forest
```{r,eval=FALSE}

library(randomForest)
rf.fit=randomForest(dep_delay∼.,data=fl_tr[-5], mtry=6,importance=TRUE,ntree=100) 

```

###GAM


```{r}

library(gam)
form <- formula(dep_delay ~ s(dep_date) + s(sched_dep_time) + carrier + origin +tzone  + s(logdistance) +
                  s(temp) + s(dewp) + s(humid) + s(wind_dir) + s(wind_speed) + precip + s(visib))
gam_fit <- gam(form, data=fl_tr,family=gaussian) 
summary(gam_fit)
plot(gam_fit,se=TRUE)
gam_pred <- predict(gam_fit,newdata=fl_te)
mse_gam <- mean((fl_te$dep_delay-gam_pred)^2)
mse_gam

```


###Gradient Boost running with default hyperparameters


```{r,eval=FALSE}

library(gbm)
dep_date_numeric <- as.numeric(fl_tr$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
fl_tr_tem <- mutate(fl_tr,dep_date = dep_date_numeric)
gbm_fit <-gbm(dep_delay ~ .,data=fl_tr_tem,distribution="gaussian",
              n.trees = 1000, shrinkage = 0.01,interaction.depth = 1)

dep_date_numeric <- as.numeric(fl_te$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
fl_te_tem <- mutate(fl_te,dep_date = dep_date_numeric)

gbm_pred <- predict(gbm_fit,newdata=fl_te_tem,n.trees = 1000)
mse_gbm <- mean((fl_te$dep_delay-gbm_pred)^2)
mse_gbm

```


##Hyper Tuning of GBM

```{r, eval=FALSE}

dvalues <- matrix(ncol=3,nrow = 4)
ntrees_list<-c(1000,2000)
list_shrink<-c(0.1,0.2)
j<-1
for(tree_i in ntrees_list)
{
  for(shrinkage_k in list_shrink){
    gbm_fit <-gbm(dep_delay ~ .,data=fl_tr_tem,distribution="gaussian",
                  n.trees = tree_i, shrinkage = shrinkage_k, interaction.depth = 3)
    
    gbm_pred <- predict(gbm_fit,newdata=fl_te_tem,n.trees =tree_i)
    mse_gbm <- mean((fl_te$dep_delay-gbm_pred)^2)
    dvalues[j,1] <- tree_i
    dvalues[j,2] <- shrinkage_k
    dvalues[j,3] <- mse_gbm
    j<-j+1
    
  }
  }
 index<-which(dvalues[,3] == min(dvalues[,3]), arr.ind = T) 
 best_tree<-dvalues[index,1]
 best_shrink<-dvalues[index,2]
```
###Analyzing mse_list, the last value of parameters, shrinkage=0.2 and trees=2000 give best result.

###Training with tuned paramaters

```{r,eval=TRUE}

library(gbm)
dep_date_numeric <- as.numeric(fl_tr$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
fl_tr_tem <- mutate(fl_tr,dep_date = dep_date_numeric)
gbm_fit <-gbm(dep_delay ~ .,data=fl_tr_tem,distribution="gaussian",
              n.trees = 2000, shrinkage = 0.2,interaction.depth = 3)

dep_date_numeric <- as.numeric(fl_te$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
fl_te_tem <- mutate(fl_te,dep_date = dep_date_numeric)

gbm_pred <- predict(gbm_fit,newdata=fl_te_tem,n.trees = 2000)
mse_gbm <- mean((fl_te$dep_delay-gbm_pred)^2)
mse_gbm

```


##Tree

```{r}

library(tree)
tree.pred<-tree(dep_delay ~., data=fl_tr_tem)
yhat=predict(tree.pred ,newdata=fl_te_tem) 
tree.test=fl_te_tem$dep_delay

```


```{r}

plot(tree.pred)
text(tree.pred,cex=.8)

```


```{r}

mse_t<-mean((yhat -tree.test)^2)
mse_t

```


##Using xgboost
```{r}

library(xgboost)
classifier = xgboost(data = data.matrix(fl_tr_tem[-2]), label = fl_tr_tem$dep_delay, nrounds =100)
xgb_pred<-predict(classifier,data.matrix(fl_te_tem[-2]))
mse_xgb<-mean((xgb_pred- fl_te_tem[,2])^2)
mse_xgb

```

##Running on Test Set

###Prepare Test Data and perform same preprocessing as while during training
###Predicting Fitted Tuned Model on Test Set
```{r}

fltest <- read.csv("C:\\Users\\Karan\\Downloads\\stat\\project\\fltest.csv.gz")
flt <- fltest
for(i in 1:ncol(fl)) {
  if(typeof(fl[[i]]) == "character") {
    flt[[i]] <- factor(flt[[i]])
  }
}

p<-function(x) {sum(is.na(x))/length(x)*100}
apply(flt,2,p)
flt <- flt%>% 
  select(-type,-manufacturer,-model,-engines,-seats, -speed, -engine,-wind_gust,-pressure,-year.y)

flt <- na.omit(flt)

dent <- nrow(flt)+1
flt <- flt %>% mutate(dep_delay = rank(dep_delay)/dent)

flt <- flt %>% 
  mutate(dep_date = make_date(year.x,month,day)) %>% 
  select(-year.x,-month,-day,-dep_time,-arr_time,-arr_delay,
         -sched_arr_time,-tailnum,-flight,-name,-air_time,
         -hour,-minute,-time_hour,-tz,-dst,-dest) %>%
  mutate(precip = as.numeric(precip>0))

flt <- mutate(flt,logdistance = log(distance)) %>% select(-distance)
flt <- mutate(flt,logalt = log(alt)) %>% select(-alt)
flt <- flt %>% mutate(dep_delay = rank(dep_delay)/dent)
dep_date_numeric <- as.numeric(flt$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
flt <- mutate(flt,dep_date = dep_date_numeric)

gbm_pred_t <- predict(gbm_fit,newdata=flt,n.trees = 2000)

mse_gbm <- mean((flt$dep_delay-gbm_pred_t)^2)
mse_gbm

```



